Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-05-01 demystify.tex
updates

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipment and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

@@

\begin{commentbox}[Vu:]
  At GMU CS, admitted PhD students have 4 years of GTA guaranteed, and also receive a stipend for the first summer.
\end{commentbox}

@@

\section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}
A common question is whether you should mention that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LGBTQ+ group (e.g., in your SOP).
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

@@

It is understandable that you might feel uncomfortable disclosing this, fearing you would lost a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academic and likely would not be tolerated by the university). In my experience, it is the opposite: \emph{many profs. actively seek out to work with students from diverse backgrounds and view diversity as an asset to their research group}.

Moreover, promoting diversity aligns with the goal of broadening participation in computing (BPC) that is strongly supported within academia (e.g., see the \href{https://plans.bpcnet.org/GeorgeMasonUniversity_ComputerScience_DepartmentalBPCPlan.pdf}{BPC plan at GMU CS}) and government funding agencies such as NSF (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}).
Also, sometimes providing this information can make it easier for faculty to apply for certain scholarships or fellowships to support you, e.g., from a private donor that has offered a fellowship only to students from certain background.

@@

\paragraph{Core CS areas} Common CS areas that you should have knowledge in are:
\begin{itemize}
  \item Programming Foundation (e.g., programming in C/C++ or Java, OOP)
  \item Discrete Math (e.g., logic, set theory, proof techniques)
  \item Data Structures and Algorithms (e.g., linked lists, trees, sorting, searching)
  \item Computer OS or Systems (e.g., memory management, file systems, processes)
\end{itemize}

In short, you \emph{do not need} to formally taking CS courses, you just need to show that you have this essential knowledge, e.g., through the mentioned ways. Many universities are well-aware that incoming graduate students might not have all the technical background, so they often have a \emph{``bridge''} courses to help students catch up.  For example, GMU has four bridge courses corresponding to the four core areas above that incoming students can take to catch up on their CS knowledge.

\begin{commentbox}[Vu:]
  I would advocate for a non-STEM student who shows that they have a strong drive for CS by studying core CS knowledge through various channels (e.g., self-study through online courses, projects, etc.).  I have seen many students with non-CS background who are very successful in CS PhD.  I also have seen many students with CS background who are not successful in CS PhD.  So it is not about your background, it is about your drive and passion for CS research.
\end{commentbox}

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, student can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.  However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{sec:your-school}).

@@

\begin{commentbox}[Vu:]
    I start my PhD with an MS, and it took me 7 years (Fall'2007-- Fall'14). I did spend half a year doing internship at Naval Research Lab and 1 year doing internship at Google
  \end{commentbox}

@@

\emph{Funding}:  In many countries funding comes from the university or the gov't. This funding often has a fixed duration (e.g., 3 or 4 years).  In the US (\autoref{sec:funding}), funding (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in European universities compared to the US.

>>>>

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipment and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

@@

\begin{commentbox}[Vu:]
  At GMU CS, admitted PhD students have 4 years of GTA guaranteed, and also receive a stipend for the first summer (\autoref{sec:summer-funding}).
\end{commentbox}

@@

\section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}

A common question is whether you should mention that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LGBTQ+ group (e.g., in your SOP).
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

@@

It is understandable that you might feel uncomfortable disclosing this, fearing you would lose a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academics and likely would not be tolerated by the university). In my experience, it is the opposite: \emph{many profs. actively seek to work with students from diverse backgrounds and view diversity as an asset to their research group}.

Moreover, promoting diversity aligns with the goal of broadening participation in computing (BPC) that is strongly supported within academia (e.g., see the \href{https://plans.bpcnet.org/GeorgeMasonUniversity_ComputerScience_DepartmentalBPCPlan.pdf}{BPC plan at GMU CS}) and government funding agencies such as NSF (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}).
Also, sometimes providing this information can make it easier for faculty to apply for certain scholarships or fellowships to support you, e.g., from a private donor that has offered a fellowship only to students from certain backgrounds.

@@

\paragraph{Core CS areas} Common CS areas that you should know are:
\begin{itemize}
  \item Programming Foundation (e.g., programming in C/C++ or Java, OOP)
  \item Discrete Math (e.g., logic, set theory, proof techniques)
  \item Data Structures and Algorithms (e.g., linked lists, trees, sorting, searching)
  \item Computer OS or Systems (e.g., memory management, file systems, processes)
\end{itemize}

In short, you \emph{do not need} to formally take CS courses, you just need to show that you have this essential knowledge, e.g., through the mentioned ways. Many universities are well aware that incoming graduate students might not have all the technical background, so they often have a \emph{``bridge''} courses to help students catch up.  For example, GMU has four bridge courses corresponding to the four core areas above that incoming students can take to catch up on their CS knowledge.

\begin{commentbox}[Vu:]
  I would advocate for a non-STEM student who shows that they have a strong drive for CS by studying core CS knowledge through various channels (e.g., self-study through online courses, projects, etc.).  I have seen many students with non-CS backgrounds who are very successful in CS PhD.  I also have seen many students with CS background who are not successful in CS PhD.  So it is not about your background, it is about your drive and passion for CS research.
\end{commentbox}

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.  However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{sec:your-school}).

@@

% \begin{commentbox}[Vu:]
%     I start my PhD with an MS, and it took me 7 years (Fall'2007-- Fall'14). I spent half a year doing an internship at the Naval Research Lab.
%   \end{commentbox}

@@

\emph{Funding}:  In many countries, funding comes from the university or the gov't. This funding often has a fixed duration (e.g., 3 or 4 years).  In the US (\autoref{sec:funding}), funding (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in European universities compared to the US.
